\addcontentsline{toc}{chapter}{Acknowledgments}
\thispagestyle{empty}

\vspace*{20mm}

\begin{center}
    {\usekomafont{sectioning}\usekomafont{section} Acknowledgments}
\end{center}

\vspace{10mm}

I would like to express my sincere gratitude to my main advisor, Prof.\ Dr.\ Martin Schreiber from the Technical University of Munich (TUM), for his continuous guidance, insightful discussions, and constructive feedback throughout the course of this thesis. His expertise in parallel programming and runtime systems, as well as his encouragement to critically analyze both practical implementations and standardization aspects, significantly shaped the direction and quality of this work.

I would also like to thank my advisors from RWTH Aachen University for their valuable input and discussions on parallel runtime systems and resource management, which contributed important perspectives to this thesis.

Furthermore, I am grateful to the members of the research group for the stimulating environment and informal exchanges that helped deepen my understanding of OpenMP runtimes and dynamic resource management.

Above all, I would like to express my deepest gratitude to my wife. Her constant support, patience, and understanding made it possible for me to pursue this work alongside a full-time job, the responsibilities of running my own company, and the completion of this master's thesis in parallel. Without her encouragement and support, this work would not have been possible.

Finally, I would like to thank my family and friends for their continued support throughout my studies.

\cleardoublepage{}
